Types for R, Empirically Designed and Evaluated

### 2020-05-07 paper/main.tex
Fix quotes

%
%
\subsubsection{Challenges}
There are two interesting challenges that need to be tackled to make this work
reliably, without surprises for the end-users. Firstly, we need to correctly
type-check arguments. R is a non-strict language~\cite{oopsla19};
when a function is called, instead of getting values as arguments which can be
immediately type-checked, we get unevaluated code thunks. We can force the
evaluation of these thunks to obtain values for type-checking but that will not
preserve the language’s semantics. This will lead to incorrect results and
surprises for the end-users. To preserve the original semantics, we modify the
thunks to call our type-checker on the value obtained when they are forced,
along with the context necessary to associate the thunk to the function and its
parameter (for failure messages). This works, but with one wrinkle. GNU R has a
bytecode compiler that can sometimes optimize away thunks, in those cases, we
don’t get thunks as arguments but values that can be immediately type-checked.
However, we still can’t type-check the value and issue failure messages because
that would violate the semantics. Non-strict semantics require that any
computation related to the argument happens on the first use of the argument. If
this argument was never used, it would be incorrect to type-check it. To make
this work correctly, we explicitly wrap the argument value in a thunk with a
call to the type-checker as in the previous case.  Secondly, we need to
correctly type-check function return values. In R, the result of function
evaluation is the result of evaluating the last expression in its body; so a
return call is absent from most function definitions. So there are many
potential subexpressions which need to be type-checked aginst the return type.
For this, we register our type-checker to be called on the return value in a
function exit hook. This hook is executed in the function’s environment after it
has finished executing. While this works, we also have to contend with yet
another wrinkle, R interpreter can perform longjumps in its C implementation,
which causes R functions on the stack to be discarded. When they are discarded,
their exit hooks are called, which in our case calls the type-checker. But,
these functions are in the middle of an active computation, so they don’t have a
return value to type-check. We deal with this problem by allocating a unique
object which serves as the return value for calls that are discarded. Our exit
hook does not call the type-checker if this unique value happens to be the
return value of the function.

@@

%
%
\subsubsection{Usability}
Through this package, we intend to provide a clear path for the R developers to
add types to untyped package code. Hence, we have paid a lot of attention to
usability in its design.
Firstly, the package enables type-checking for all packages with type
signatures, automatically, on importing, offering a seamless and hassle-free
experience for both package developers and end-users. A mere library(contractr)
is enough to insert contracts, both, in packages pre-loaded in the user’s
workspace and in packages that will be loaded eventually. For this, contractr
relies heavily on R’s reflective and dynamic capabilities. Upon loading
contractr scans all the pre-loaded packages in the user’s workspace and inserts
contracts in functions for which type signatures are available. For all other
packages installed on user’s machine but not yet loaded in the workspace,
contractr sets up package load hooks. The hooks get executed when these packages
get loaded and insert contracts to the package functions. The ensures that users
do not fiddle manually with the contractr API to enable type-checking for
packages loaded eventually. Furthermore, contractr automatically removes
contracts from all the package functions and restores them to their original
state when it is unloaded, an operation that is rarely performed but is
possible, nonetheless.
Secondly, we enable package authors to supply type declarations to contractr for
their package functions, without requiring any modification to package code.
This is accomplished by providing type declarations inside roxygen2 function
comment blocks. roxygen2 is an R package that enables authors to add
documentation to R functions in the form of organized plaintext sections which
are automatically exported to R’s latex style custom documentation format.
roxygen2 is used by … R packages in CRAN. It also provides an extension API for
processing metadata in custom documentation tag sections. A package can register
a method to be called when this tag is encountered. The metadata for that tag
section is made available to the registrant before package build and
installation steps. contractr uses this API to implement a @type tag. Our
roxygen2 hook extracts type declarations from these sections from all the
documented functions of the package and stores then in a TYPEDECLARATION file
inside the package folder. This files is copied as is on package installation
and automatically picked up contractr when the two packges are loaded.
Here is an example of how a roxygen code block looks like with our type declaration:

>>>>

%
%
\subsubsection{Challenges}
There are two interesting challenges that need to be tackled to make this work
reliably, without surprises for the end-users. Firstly, we need to correctly
type-check arguments. R is a non-strict language~\cite{oopsla19};
when a function is called, instead of getting values as arguments which can be
immediately type-checked, we get unevaluated code thunks. We can force the
evaluation of these thunks to obtain values for type-checking but that will not
preserve the language's semantics. This will lead to incorrect results and
surprises for the end-users. To preserve the original semantics, we modify the
thunks to call our type-checker on the value obtained when they are forced,
along with the context necessary to associate the thunk to the function and its
parameter (for failure messages). This works, but with one wrinkle. GNU R has a
bytecode compiler that can sometimes optimize away thunks, in those cases, we
don't get thunks as arguments but values that can be immediately type-checked.
However, we still can't type-check the value and issue failure messages because
that would violate the semantics. Non-strict semantics require that any
computation related to the argument happens on the first use of the argument. If
this argument was never used, it would be incorrect to type-check it. To make
this work correctly, we explicitly wrap the argument value in a thunk with a
call to the type-checker as in the previous case.  Secondly, we need to
correctly type-check function return values. In R, the result of function
evaluation is the result of evaluating the last expression in its body; so a
return call is absent from most function definitions. So there are many
potential subexpressions which need to be type-checked aginst the return type.
For this, we register our type-checker to be called on the return value in a
function exit hook. This hook is executed in the function's environment after it
has finished executing. While this works, we also have to contend with yet
another wrinkle, R interpreter can perform longjumps in its C implementation,
which causes R functions on the stack to be discarded. When they are discarded,
their exit hooks are called, which in our case calls the type-checker. But,
these functions are in the middle of an active computation, so they don't have a
return value to type-check. We deal with this problem by allocating a unique
object which serves as the return value for calls that are discarded. Our exit
hook does not call the type-checker if this unique value happens to be the
return value of the function.

@@

%
%
\subsubsection{Usability}
Through this package, we intend to provide a clear path for the R developers to
add types to untyped package code. Hence, we have paid a lot of attention to
usability in its design.
Firstly, the package enables type-checking for all packages with type
signatures, automatically, on importing, offering a seamless and hassle-free
experience for both package developers and end-users. A mere library(contractr)
is enough to insert contracts, both, in packages pre-loaded in the user's
workspace and in packages that will be loaded eventually. For this, contractr
relies heavily on R's reflective and dynamic capabilities. Upon loading
contractr scans all the pre-loaded packages in the user's workspace and inserts
contracts in functions for which type signatures are available. For all other
packages installed on user's machine but not yet loaded in the workspace,
contractr sets up package load hooks. The hooks get executed when these packages
get loaded and insert contracts to the package functions. The ensures that users
do not fiddle manually with the contractr API to enable type-checking for
packages loaded eventually. Furthermore, contractr automatically removes
contracts from all the package functions and restores them to their original
state when it is unloaded, an operation that is rarely performed but is
possible, nonetheless.
Secondly, we enable package authors to supply type declarations to contractr for
their package functions, without requiring any modification to package code.
This is accomplished by providing type declarations inside roxygen2 function
comment blocks. roxygen2 is an R package that enables authors to add
documentation to R functions in the form of organized plaintext sections which
are automatically exported to R's latex style custom documentation format.
roxygen2 is used by … R packages in CRAN. It also provides an extension API for
processing metadata in custom documentation tag sections. A package can register
a method to be called when this tag is encountered. The metadata for that tag
section is made available to the registrant before package build and
installation steps. contractr uses this API to implement a @type tag. Our
roxygen2 hook extracts type declarations from these sections from all the
documented functions of the package and stores then in a TYPEDECLARATION file
inside the package folder. This files is copied as is on package installation
and automatically picked up contractr when the two packges are loaded.
Here is an example of how a roxygen code block looks like with our type declaration:
